Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-08 paper/paper.tex
Continuing with path graphs and rewriting old text.

\subjclass{E.1 Data Structures}
\keywords{Dummy keyword -- please provide 1--5 keywords}% FIXME
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Shorthands
\newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
\newcommand{\orderk}[1]{order\nobreakdash-$#1$}
\newcommand{\LFmapping}{LF\nobreakdash-mapping}
\newcommand{\FMindex}{FM\nobreakdash-index}

@@

\subsection{Strings}

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively.

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$, where set $Z$ is one of $\mathbb{Z}$, $\set{x \in \mathbb{Z} \mid x \ge 0}$, and $\set{x \in \mathbb{Z} \mid x \le 0}$. The labels of such paths are infinite sequences of characters instead of strings.

@@

% FIXME example of strings, DBG, long paths, false positives
Let $\mathcal{S}$ be a collection of strings over alphabet $\Sigma$. We assume that each string $S \in \mathcal{S}$ begins with $k$ copies of $\#$ and ends with $k$ copies of $\$$, with no other occurrences of these characters in the string. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$. Each node $v_{X}$ has a \emph{key} $G.\gkey(v_{X}) = X$ and a label $G.\glabel(v_{X}) = X[0]$. The edges of the de~Bruijn graph represent \kmer{k+1}s $Y$ occurring in the collection. For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$. We use the node corresponding to substring $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.

\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ containing at most one $\$$. Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
\end{lemma}

\begin{proof}
Let $X = S[i, i+\abs{X}-1]$ for a string $S \in \mathcal{S}$. For $i \le j \le i+\abs{X}-1$, there is node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$. By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$. Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then the label of the path is a substring of a string $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$. If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$. Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

@@

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. Before proceeding, we first have to define the de~Bruijn graph of a labeled graph. Let $P$ be a (finite or infinite) path in graph $G = (V, E)$. We say that path $P$ is \emph{maximal}, if starts at the source node or extends infinitely to the left, and ends at the sink node or extends infinitely to the right. We construct a collection of character sequences $\mathcal{S}$ by taking all maximal paths $P$ in graph $G$ and creating a sequence $S$ based on $G.\glabel(P)$ for each such path. If path $P$ is infinite in both directions, we use the label directly. If the path starts at the source node, we adjust the number of characters $\#$ at the beginning to match the order $k$ of the de~Bruijn graph. Similarly, if the path ends at the sink node, we adjust the number of characters $\$$ at the end. We can then generalize the definition of de~Bruijn graphs in Section~\ref{sect:graphs} to create a de~Bruijn graph of sequence collection $\mathcal{S}$.

@@

% FIXME from here
using de Bruijn graphs as indexes; false positives
- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
- valid for existence queries of length up to k+1 and locate queries of length up to k

Path graphs
- a generalization of DBG as a path index
- keys must be prefix-free
- k-index: has a node matching any k-path in the input graph
- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
- k-equivalent: union of values for nodes matching X identical for all |X| <= k
- k-equivalent + edge condition: no false positives of length <= k; no false negatives

@@

>>>>

\subjclass{E.1 Data Structures}
\keywords{Burrows-Wheeler transform, de Bruijn graphs, path indexes, variation graphs}
% Author macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

% Shorthands
\newcommand{\kmer}[1]{$#1$\nobreakdash-mer}
\newcommand{\kcollection}[1]{$#1$\nobreakdash-collection}
\newcommand{\orderk}[1]{order\nobreakdash-$#1$}
\newcommand{\LFmapping}{LF\nobreakdash-mapping}
\newcommand{\FMindex}{FM\nobreakdash-index}

@@

\subsection{Strings}\label{sect:strings}

A \emph{string} $S[0, n-1] = s_{0} \dotsm s_{n-1}$ of length $\abs{S} = n$ is a sequence of \emph{characters} over an \emph{alphabet} $\Sigma = \set{0, \dotsc, \sigma - 1}$. For indexing purposes, we often consider \emph{text} strings $T[0, n-1]$ terminated by an \emph{endmarker} $T[n-1] = \$ = 0$ not found anywhere else in the text. \emph{Binary} sequences are strings over the alphabet $\set{0, 1}$. A \emph{substring} of string $S$ is a sequence of the form $S[i, j] = s_{i} \dotsm s_{j}$. We call substrings of the type $S[0, j]$ and $S[i, n-1]$ \emph{prefixes} and \emph{suffixes}, respectively, and substrings of length $k$ as \kmer{k}s. We say that string $S'$ is a substring of string collection $\mathcal{S}$, if there is a string $S \in \mathcal{S}$ such that $S'$ is a substring of $S$.

In some cases, we also consider potentially \emph{infinite} character sequences $S = (s_{i})_{i \in Z}$, where set $Z$ is a contiguous subset of $\mathbb{Z}$. The notion of substrings generalizes for such infinite sequences in a natural way. We say that a substring of an infinite sequence $S$ is \emph{left-infinite} if it extends infinitely to the left, and \emph{right-infinite} if it extends infinitely to the right. A substring of a finite or infinite sequence $S$ is \emph{left-maximal} if it is left-infinite or a prefix; \emph{right-maximal} if it is right-infinite or a suffix; and \emph{maximal} if it is both left-maximal and right-maximal.

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$ as we did with infinite character sequences in Section~\ref{sect:strings}.

@@

\begin{definition}[\kcollection{k}]
Let $\mathcal{S}$ be a collection of character sequences over alphabet $\Sigma$, and let $k$ be a parameter value. We say that $\mathcal{S}$ is a \emph{\kcollection{k}}, if each sequence $S \in \mathcal{S}$
\begin{enumerate}
\item is left-infinite or begins with $\#^{k}$;
\item is right-infinite or ends with $\$^{k}$; and
\item contains no other occurrences of characters $\#$ and $\$$,
\end{enumerate}
and the collection contains substrings $\#^{k}$ and $\$^{k}$.
\end{definition}

% FIXME example of strings, DBG, long paths, false positives
% FIXME some comments about DBGs
\begin{definition}[de~Bruijn graph]
Let $\mathcal{S}$ be a \kcollection{k} over alphabet $\Sigma$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$ such that
\begin{enumerate}
\item each node $v_{X} \in V$ represent a distinct \kmer{k} occurring in $\mathcal{S}$;
\item $G.\glabel(v_{X}) = X[0]$ and $G.\gkey(v_{X}) = X$ for all $v_{X} \in V$; and
\item each edge $(v_{X}, v_{X'}) \in E$ represents a distinct \kmer{k+1} $Y = Xc = c'X'$ (where $c, c' \in \Sigma$) occurring in $\mathcal{S}$.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $X$ be a finite substring of $\mathcal{S}$ containing at most one $\$$. Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
\end{lemma}

\begin{proof}
Let $X = S[i, i+\abs{X}-1]$ for a sequence $S \in \mathcal{S}$. For $i \le j \le i+\abs{X}-1$, there is a node $v_{j} \in V$ representing substring $X_{j} = S[j, j+k-1]$, with $G.\glabel(v_{j}) = S[j]$. By definition, the de~Bruijn graph has an edge $(v_{j}, v_{j+1})$ for $i \le j < i+\abs{X}-1$. Hence $P = v_{i} \dotsm v_{i+\abs{X}-1}$ is a path in $G$ and $G.\glabel(P) = X$.
\end{proof}

\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of \kcollection{k} $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then the label of the path is a substring of a sequence $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
If $\abs{P} = 0$, the label is an empty string and hence a substring of each sequence $S \in \mathcal{S}$.
If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.
Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

@@

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. We now define the de~Bruijn graph of a graph for that purpose.

\begin{definition}[Maximal paths]
Let $G = (V, E)$ be a labeled graph. We say that path $P$ in the graph is \emph{left-maximal} if it starts at the source node or extends infinitely to the left; \emph{right-maximal} if it ends at the sink node or extends infinitely to the right; and \emph{maximal} if it is both left-maximal and right-maximal.
\end{definition}

We can build a \kcollection{k} of maximal path labels by starting with the collection $\mathcal{S}$ of maximal path labels. In order to transform $\mathcal{S}$ into a \kcollection{k}, we adjust the number of characters $\#$ at the beginning of each non-left-infinite sequence, and the number of characters $\$$ at the end of each non-right-infinite sequence. We also add string $\#^{k} \$^{k}$ into the collection if necessary.

\begin{definition}[de~Bruijn graph of a graph]
Let $G = (V, E)$ be a labeled graph, and let $\mathcal{S}$ be a \kcollection{k} of maximal path labels in $G$. The \orderk{k} de~Bruijn graph of $\mathcal{S}$ is also the \orderk{k} de~Bruijn graph of graph $G$.
\end{definition}

@@

De~Bruijn graphs are a special case of path graphs. They represent paths of length $k$ in the \emph{input graph} as nodes and paths overlapping on $k-1$ nodes of the input graph as edges. We can generalize the notion by allowing the nodes of the path graph to represent paths of different lengths.

\begin{definition}[Prefix-matching strings]
Let $S$ and $S'$ be strings over alphabet $\Sigma$. We say that string $S$ and $S'$ \emph{prefix-match}, if $S$ is a prefix of $S'$ or $S'$ is a prefix of $S$.
\end{definition}

\begin{definition}[Prefix-free set]
Let $\mathcal{S}$ be a set of strings over alphabet $\Sigma$. We say that set $\mathcal{S}$ is \emph{prefix-free}, if no two strings $S, S' \in \mathcal{S}$ (with $S \ne S'$) prefix-match.
\end{definition}

\begin{definition}[Path graph]
Let $\mathcal{S}$ be a \kcollection{k} of the labels of maximal paths in graph $G = (V, E)$, and let $\mathcal{K}$ be a prefix-free set of strings containing $\#^{k}$ and $\$^{k}$. We assume that each right-maximal substring $S$ in $\mathcal{S}$ prefix-matches a string $K \in \mathcal{K}$, and that each string $K \in \mathcal{K}$ is a substring of $\mathcal{S}$. Then the \orderk{k} \emph{path graph} of graph $G$ with \emph{key set} $\mathcal{K}$ is a graph $G' = (V', E')$ such that
\begin{enumerate}
\item each node $v_{K} \in V'$ represent a distinct key $K \in \mathcal{K}$;
\item $G'.\glabel(v_{K}) = K[0]$ and $G'.\gkey(v_{K}) = K$ for all $v_{K} \in V$; and
\item each edge $(v_{K}, v_{K'}) \in E'$ represents the occurrence of substring $K[0] K'$ in $\mathcal{S}$ such that strings $K$ and $K[0] K'$ prefix-match.
\end{enumerate}
We use the node corresponding to $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.
\end{definition}

% FIXME from here
- k-index: has a node matching any k-path in the input graph
- edge condition: Y has predecessor with c -> edge (X, Y) if cY matches X
- k-equivalent: union of values for nodes matching X identical for all |X| <= k
- k-equivalent + edge condition: no false positives of length <= k; no false negatives

@@

using pruned de Bruijn graphs as indexes; false positives
- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
- valid for existence queries of length up to k+1 and locate queries of length up to k
